---
output:
  md_document:
    variant: markdown_github
---

```{r echo=FALSE, results='asis'}
## Insert 'inspiration', 'exploration', or 'delivery'
envreportutils::devex_badge('exploration')
```


<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# tidyhydat

Here is a list of what tidyhydat does:
- Perform a number of common queries on the HYDAT database and returns a tibble
- Maintains column names as the database itself
- Can select one, two... x stations
- Keep functions are low-level as possible. For example, for daily flows, the function should query the database then format the dates and that is it. 

## Installation
To install the tidyhydat package, you need to install the devtools package then the tidyhydat package
```{r, echo=TRUE, eval=FALSE}
install.packages("devtools")
devtools::install_github("bcgov/tidyhydat")
```
Then to load the package you need to use the library command:
```{r, eval = TRUE, echo=TRUE}
library(tidyhydat)
```


## Example

This is a basic example of `tidyhydat` usage

```{r example}
DLY_FLOWS(STATION_NUMBER = c("08LA001","08LG006"))
```

## Basin realtime data acquisition usage
Using `download_realtime()` we can easily select specific stations by supplying a station of interest:
```{r}
download_realtime(STATION_NUMBER = "08LG006")
```

## Example with spatial data
The `download_realtime()` functions allows us to directly query the Environment Canada and Climate Change datamart selecting by station. If we wanted to look at all the realtime stations in a particular hydrologic zone, we could easily do this using the `dplyr`,`sf` and `bcmaps` packages. To install those packages use these commands: 
```{r, echo=TRUE, eval=FALSE}
devtools::install_github("bcgov/bcmaps")
install.packages(c("sf","dplyr")
```

And then load these packages. tidyhydat is already loaded above.
```{r, eval = TRUE, echo=TRUE}
library(bcmaps)
library(sf)
library(dplyr)


```
Now to return the question. BC is divided into hydrologic zones. If use the hydrozones layer in `bcmaps` and convert it to `sf` format, determining which stations reside in which hydrologic zone is trivial. Using `st_join` allows to ask which hydrometric stations (called by `download_network`) are in which hydrologic zones. If we are interested in all stations in the QUEEN CHARLOTTE ISLANDS hydrologic zone, we can generate that list by filtering by the relevant hydrologic zone:
```{r, eval = TRUE, echo=TRUE}
## Convert to sf format
hydrozones_sf <- st_as_sf(bcmaps::hydrozones) %>%
  select(HYDZN_NAME)


qci_stations <- download_network(PROV_TERR_STATE_LOC = "BC") %>%
  st_as_sf(., coords = c("LONGITUDE", "LATITUDE"), 
              crs = 4326, 
              agr = "constant") %>%
  st_transform(crs = 3005) %>%
  st_join(.,hydrozones_sf) %>%
  filter(HYDZN_NAME == "QUEEN CHARLOTTE ISLANDS") %>%
  pull(STATION_NUMBER)

qci_stations
```

Now that vector (`qci_stations`) is useful to select which stations we are interested in. 
```{r}
qci_realtime <- download_realtime(STATION_NUMBER = qci_stations)
```

Then using `ggplot2` we could plot these results to have look at the data
```{r}
library(ggplot2)

ggplot(qci_realtime, aes(x = date_time, y = FLOW)) +
  geom_line(aes(colour = STATION_NUMBER))

```
