---
output:
  md_document:
    variant: markdown_github
---

```{r echo=FALSE, results='asis'}
## Insert 'inspiration', 'exploration', or 'delivery'
envreportutils::devex_badge('exploration')
```


<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# tidyhydat

Here is a list of what tidyhydat does:

- Perform a number of common queries on the HYDAT database and returns a tibble
- Maintains the same column names as the database itself
- Can select one, two... x stations
- Can provide a date range
- Keep functions are low-level as possible. For example, for daily flows, the function should query the database then format the dates and that is it. 
- An additional auxiliary feature outside the HYDAT database is the downloading of realtime data. This functionality is provided by `download_realtime()`, `download_network()` and `download_ws`. 

## Installation
To install the `tidyhydat` package, you need to install the `devtools` package then the `tidyhydat` package
```{r, echo=TRUE, eval=FALSE}
install.packages("devtools")
devtools::install_github("bcgov/tidyhydat")
```
Then to load the package you need to use the `library` function. When you install `tidyhydat`, several other packages will be installed as well. One of those packages, `dplyr`, is useful for data manipulations and is used regularly here. Even though `dplyr` is installed alongside `tidyhydat`, it is helpful to load it by itself as there are many useful functions contained within `dplyr`. A helpful `dplyr` tutorial can be found [here](https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html).
```{r, eval = TRUE, echo=TRUE}
library(tidyhydat)
library(dplyr)
```

## HYDAT download
To use most of the `tidyhydat` package you will need the most recent version of the HYDAT database. The sqlite3 version can be downloaded here:

- http://collaboration.cmc.ec.gc.ca/cmc/hydrometrics/www/

You will need to download that file, unzip it and put it somewhere on local storage. The path to the sqlite3 must be specified within each function that uses HYDAT. If the path is very long it may be useful to store it as an object with a smaller name like this:
```{r eval = FALSE}
hydat_loc = "A very long path that points to the HYDAT sqlite3 database"
```

These are the functions that are currently implemented to query HYDAT:

- `STATIONS`
- `DLY_FLOWS`
- `DLY_LEVELS`
- `ANNUAL_STATISTICS`

This is a list of function to be implemented:

- `ANNUAL_INSTANT_PEAKS`
- `SED_DLY_LOADS`
- `SED_DLY_SUSCON`
- `SED_SAMPLES_PSD`

## Example

This is a basic example of `tidyhydat` usage. All functions that interact with HYDAT are capitalized (e.g. `STATIONS`). These functions follow a common argument structure which can be illustrated with the `DLY_FLOWS()` function. If you would like to extract only station `08LA001` you must supply the `STATION_NUMBER` and the `PROV_TERR_STATE_LOC` arguments. The `hydat_path` argument is supplied here as a local path the database. Yours will be different. 

```{r example, warning=FALSE}
DLY_FLOWS(STATION_NUMBER = "08LA001", PROV_TERR_STATE_LOC = "BC", hydat_path = "H:/Hydat.sqlite3")
```

If you would like to query the database for two or more stations you would combine the `STATION_NUMBER` into a vector using `c()`:
```{r warning=FALSE}
DLY_FLOWS(STATION_NUMBER = c("08LA001","08NL071"), PROV_TERR_STATE_LOC = "BC", hydat_path = "H:/Hydat.sqlite3")
```

If instead you would like to extract all stations from a jurisdictions, you can use the "ALL" argument for `STATION_NUMBER`:
```{r warning=FALSE}
DLY_FLOWS(STATION_NUMBER = "ALL", PROV_TERR_STATE_LOC = "PE", hydat_path = "H:/Hydat.sqlite3")
```

In all the previous examples, no start or end date was specified. In those cases all dates in HYDAT were returned. If, however, we were only interested in a subset of dates we could use the `start_date` and `end_date` arguments. A date must be supplied to both these arguments in the form of YYYY-MM-DD. If you were interested in all daily flow data from station number "08LA001" for 1981, you would specify all days in 1981 :
```{r warning=FALSE}
DLY_FLOWS(STATION_NUMBER = "08LA001", PROV_TERR_STATE_LOC = "BC", hydat_path = "H:/Hydat.sqlite3",
          start_date = "1981-01-01", end_date = "1981-12-31")
```

## Basin realtime data acquisition usage
To download realtime data we can use approximately the same conventions discussed above. All non-HYDAT functions are in lower case. Using `download_realtime()` we can easily select specific stations by supplying a station of interest. Note that again, we need to supply both the station and the province that we are interested in but do not need to supply the HYDAT location because we accessing this data online:
```{r}
download_realtime(STATION_NUMBER = "08LG006", PROV_TERR_STATE_LOC = "BC")
```

## Downloading by jurisdiction
We can use the `download_network()` functionality to get a vector of stations by jurisdiction. For example, we can choose all the stations in Prince Edward Island using the following:
```{r}
download_realtime(STATION_NUMBER = "ALL", PROV_TERR_STATE_LOC = "PE")
```

## Web service realtime data 
We can also download realtime data using the ECCC web service via the `download_ws()` function. To retrieve data in this manner, we must do a two stage process whereby we get a token from the webservice then use that token to access the web service. Credentials to get the token are also required and can only be requested from ECCC:

```{r, eval=FALSE}
## Get token
token_out <- get_ws_token(username = username, password = password)

## Input STATION_NUMBER, parameters and date range
ws_test <- download_ws_realtime(STATION_NUMBER = "08LG006",
                                parameters = c(46,5), ## see data("param_id") for a list of codes
                                start_date = "2017-06-25",
                                end_date = "2017-07-24",
                                token = token_out)
```

## Compare download_ws and download_realtime
`tidyhydat` provides two methods to download realtime data. `download_realtime()` provides a function to import openly accessible .csv files from [here](http://dd.weather.gc.ca/hydrometric/csv/BC/). `download_ws()`, coupled with `get_ws_token()`, is an API client for a web service hosted by ECCC. `download_ws()` has several difference to `download_realtime()`. These include:

- *Speed*: The `download_ws()` is much faster for larger queries (i.e. many stations). For single station queries `download_realtime()` if more appropriate.
- *Length of record*: `download_ws()` records goes back further though only two months of data can accessed at one time. 
- *Type of parameters*: `download_realtime()` are retricted to river flow (either LEVEL and FLOW) data. In contrast `download_ws()` can download several different parameters depending on what is available for that station. See `data("param_id")` for a list and explanation of the parameters.
- *Date/Time filtering*: `download_ws()` provides argument to select a date range. Selecting a data range with `download_realtime()` is not possible until after all files have been downloaded. 
- *Accessibility*: `download_realtime()` downloads data that openly accessible. `download_ws()` downloads data using a username and password which must be provided by ECCC. 

### On the distinction between STATIONS() and download_network()
`STATIONS()` and `download_network()` perform similar tasks albeit on different data sources. `STATIONS()` extracts directly from the HYDAT sqlite3 database. In addition to realtime stations, `STATIONS()` outputs discontinued and non-realtime stations:
```{r stations}
STATIONS(STATION_NUMBER = "ALL", PROV_TERR_STATE_LOC = "PE", hydat_path = "H:/Hydat.sqlite3")
```

This is contrast to `download_network()` which downloads all realtime stations. Though this is not always the case, it is best to use `download_network()` when dealing with realtime data and `STATIONS()` when interacting with HYDAT.  

## Example with spatial data
The `download_realtime()` functions allows us to directly query the Environment Canada and Climate Change datamart selecting by station. If we wanted to look at all the realtime stations in a particular hydrologic zone, we could easily do this using the `dplyr`,`sf` and `bcmaps` packages. To install those packages use these commands: 
```{r, echo=TRUE, eval=FALSE}
devtools::install_github("bcgov/bcmaps")
install.packages("sf")
```

And then load these packages. `tidyhydat` and `dplyr` is already loaded above.
```{r, eval = TRUE, echo=TRUE}
library(bcmaps)
library(sf)
```
Now to return the question. BC is divided into hydrologic zones. If use the hydrozones layer in `bcmaps` and convert it to `sf` format, determining which stations reside in which hydrologic zone is trivial. Using `st_join` allows to ask which hydrometric stations in the realtime network (called by `download_network`) are in which hydrologic zones. If we are interested in all realtime stations in the QUEEN CHARLOTTE ISLANDS hydrologic zone, we can generate that list by filtering by the relevant hydrologic zone:
```{r, eval = TRUE, echo=TRUE}
## Convert to sf format
hydrozones_sf <- st_as_sf(bcmaps::hydrozones) %>%
  select(HYDZN_NAME)


qci_stations <- download_network(PROV_TERR_STATE_LOC = "BC") %>%
  st_as_sf(., coords = c("LONGITUDE", "LATITUDE"), 
              crs = 4326, 
              agr = "constant") %>%
  st_transform(crs = 3005) %>%
  st_join(.,hydrozones_sf) %>%
  filter(HYDZN_NAME == "QUEEN CHARLOTTE ISLANDS") %>%
  pull(STATION_NUMBER)

qci_stations
```

Now that vector (`qci_stations`) is useful to select which stations we are interested in. 
```{r}
qci_realtime <- download_realtime(STATION_NUMBER = qci_stations, PROV_TERR_STATE_LOC = "BC")
```

Then using `ggplot2` we could plot these results to have look at the data
```{r}
library(ggplot2)

ggplot(qci_realtime, aes(x = date_time, y = FLOW)) +
  geom_line(aes(colour = STATION_NUMBER))
```

## Project Status

This package is under continual development.

## Getting Help or Reporting an Issue

To report bugs/issues/feature requests, please file an [issue](https://github.com/bcgov/tidyhydat/issues/).

## How to Contribute

If you would like to contribute to the package, please see our 
[CONTRIBUTING](CONTRIBUTING.md) guidelines.

Please note that this project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md). By participating in this project you agree to abide by its terms.

## License

    Copyright 2015 Province of British Columbia

    Licensed under the Apache License, Version 2.0 (the "License");
    you may not use this file except in compliance with the License.
    You may obtain a copy of the License at 

       http://www.apache.org/licenses/LICENSE-2.0

    Unless required by applicable law or agreed to in writing, software
    distributed under the License is distributed on an "AS IS" BASIS,
    WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
    See the License for the specific language governing permissions and
    limitations under the License.
