---
output:
  md_document:
    variant: markdown_github
---

<div id="devex-badge"><a rel="Delivery" href="https://github.com/BCDevExchange/docs/blob/master/discussion/projectstates.md"><img alt="In production, but maybe in Alpha or Beta. Intended to persist and be supported." style="border-width:0" src="https://assets.bcdevexchange.org/images/badges/delivery.svg" title="In production, but maybe in Alpha or Beta. Intended to persist and be supported." /></a></div> [![Travis-CI Build Status](https://travis-ci.org/bcgov/tidyhydat.svg?branch=master)](https://travis-ci.org/bcgov/tidyhydat)

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# tidyhydat
Here is a list of what `tidyhydat` does:

- Perform a number of common queries on the HYDAT database and returns tidy data
- Keep functions are low-level as possible. For example, for daily flows, the `DLY_FLOWS()` function queries the database, *tidies* the data and returns the data. 

A more thorough vignette outlining the full functionality of `tidyhydat` is outlined [here](https://github.com/bcgov/tidyhydat/blob/master/vignettes/tidyhydat.md)

## Installation
To install the `tidyhydat` package, you need to install the `devtools` package then the `tidyhydat` package
```{r, echo=TRUE, eval=FALSE}
install.packages("devtools")
devtools::install_github("bcgov/tidyhydat")
```

Then to load the package you need to using the `library` function. When you install `tidyhydat`, several other packages will be installed as well. One of those packages, `dplyr`, is useful for data manipulations and is used regularly here. Even though `dplyr` is installed alongside `tidyhydat`, it is helpful to load it by itself as there are many useful functions contained within `dplyr`. A helpful `dplyr` tutorial can be found [here](https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html).

```{r, eval = TRUE, echo=TRUE, message=FALSE}
library(tidyhydat)
library(dplyr)
```

## HYDAT download
To use most of the `tidyhydat` package you will need the most recent version of the HYDAT database. The sqlite3 version can be downloaded here:

- http://collaboration.cmc.ec.gc.ca/cmc/hydrometrics/www/

`tidyhydat` also provides a convenience function to download hydat (be patient though this takes a long time!):
```{r, eval=FALSE}
download_hydat(hydat_path = "H:/")
```

You will need to download that file, unzip it and put it somewhere on local storage. The path to the sqlite3 must be specified within each function that uses HYDAT. If the path is very long it may be useful to store it as an object with a smaller name like this:
```{r eval = FALSE}
hydat_loc = "A very long path that points to the HYDAT sqlite3 database"
```

It will quickly get tiring manually entering `hydat_path`. R provides a means setting a hydat path once in the `.Renviron` file which is then automatically called by each HYDAT function. In RStudio you can open up `.Renviron` like this:
```{r, eval= FALSE, echo = TRUE}
file.edit("~/.Renviron")
```
This opens your `.Renviron` file which is most likely blank. Add this line:
```{r, eval=FALSE, echo=TRUE}
hydat = "YOUR HYDAT PATH"
```
It is important that you name the variable `hydat` as that is name of the variable that the functions are looking for. 

## Usage

### HYDAT
This is a basic example of `tidyhydat` usage. All functions that interact with HYDAT are capitalized (e.g. `STATIONS`). These functions follow a common argument structure which can be illustrated with the `DLY_FLOWS()` function. If you would like to extract only station `08LA001` you can supply the `STATION_NUMBER`. The `hydat_path` argument is supplied here as a local path the database. Yours will be different or missing altogether if you set the variable in your `.Renviron` file which is highly recommend. 

```{r example1, warning=FALSE}
DLY_FLOWS(hydat_path = "H:/Hydat.sqlite3", STATION_NUMBER = "08LA001")
```

If you would instead prefer all stations from a province, you can use the `PROV_TERR_STATE_LOC` argument and omit the `STATION_NUMBER` argument:

```{r example2, warning=FALSE}
DLY_FLOWS(hydat_path = "H:/Hydat.sqlite3", PROV_TERR_STATE_LOC = "PE")
```

You can also make use of auxiliary function in `tidyhdyat` called `search_name` to look for matches when you know part of a name of a station. For example:
```{r, echo=TRUE}
search_name("liard")
```

### Real-time
To download real-time data using the datamart we can use approximately the same conventions discussed above. Using `download_realtime_dd()` we can easily select specific stations by supplying a station of interest:
```{r}
download_realtime_dd(STATION_NUMBER = "08LG006")
```
Another option is to provide simply the province as an argument and download all stations from that province:
```{r}
download_realtime_dd(PROV_TERR_STATE_LOC = "PE")
```
Additionally `download_realtime_ws()` provides another means of acquiring real time data.


## Project Status

This package is under continual development.

## Getting Help or Reporting an Issue

To report bugs/issues/feature requests, please file an [issue](https://github.com/bcgov/tidyhydat/issues/).

These are very welcome!

## How to Contribute

If you would like to contribute to the package, please see our 
[CONTRIBUTING](CONTRIBUTING.md) guidelines.

Please note that this project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md). By participating in this project you agree to abide by its terms.

## License

    Copyright 2017 Province of British Columbia

    Licensed under the Apache License, Version 2.0 (the "License");
    you may not use this file except in compliance with the License.
    You may obtain a copy of the License at 

       http://www.apache.org/licenses/LICENSE-2.0

    Unless required by applicable law or agreed to in writing, software
    distributed under the License is distributed on an "AS IS" BASIS,
    WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
    See the License for the specific language governing permissions and
    limitations under the License.
