---
output:
  md_document:
    variant: markdown_github
---

```{r echo=FALSE, results='asis'}
## Insert 'inspiration', 'exploration', or 'delivery'
envreportutils::devex_badge('exploration')
```


<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# tidyhydat

Here is a list of what tidyhydat does:
- Perform a number of common queries on the HYDAT database and returns a tibble
- Maintains the same column names as the database itself
- Can select one, two... x stations
- Keep functions are low-level as possible. For example, for daily flows, the function should query the database then format the dates and that is it. 
- An additional auxiliary feature outside the HYDAT database is the downloading of realtime data. This functionality is provided by `download_realtime()` and `download_network()`. 

## Installation
To install the tidyhydat package, you need to install the devtools package then the tidyhydat package
```{r, echo=TRUE, eval=FALSE}
install.packages("devtools")
devtools::install_github("bcgov/tidyhydat")
```
Then to load the package you need to use the library command. When you install tidyhydat, several other packages will be installed as well. One of those packages, `dplyr`, is useful for data manipulations and is used regularly here. Even though `dplyr` is installed alongside `tidyhydat`, you must still load it explicitly.
```{r, eval = TRUE, echo=TRUE}
library(tidyhydat)
library(dplyr)
```

## HYDAT download
To use most of the `tidyhydat` package you will need the most recent version of the HYDAT database. The sqlite3 version can be downloaded here:

- http://collaboration.cmc.ec.gc.ca/cmc/hydrometrics/www/

You will need to download that file, unzip it and put it somewhere on local storage. The path to the sqlite3 must be specified within each function that uses HYDAT.  

## Example

This is a basic example of `tidyhydat` usage. All functions that interact with HYDAT are capitalized (e.g. STATIONS). These functions follow a common argument structure which can be illustrated with the `DLY_FLOWS()` function. If you would like to extract only station `08LA001` you must supply the `STATION_NUMBER` and the `PROV_TERR_STATE_LOC` arguments: 

```{r example, warning=FALSE}
DLY_FLOWS(STATION_NUMBER = "08LA001", PROV_TERR_STATE_LOC = "BC")
```

If you would like to query the database for two or more stations you would supply this command:
```{r warning=FALSE}
DLY_FLOWS(STATION_NUMBER = c("08LA001","08NL071"), PROV_TERR_STATE_LOC = "BC")
```

If instead you would like to download ALL stations from a jurisdictions, you can use the "ALL" argument for `STATION_NUMBER`:
```{r warning=FALSE}
DLY_FLOWS(STATION_NUMBER = "ALL", PROV_TERR_STATE_LOC = "PE")
```

## Basin realtime data acquisition usage
To download realtime data we can use approximately the same conventions discussed above. All non-HYDAT functions are in lower case. Using `download_realtime()` we can easily select specific stations by supplying a station of interest. Note that again, we need to supply both the station and the province that we are interested in:
```{r}
download_realtime(STATION_NUMBER = "08LG006", PROV_TERR_STATE_LOC = "BC")
```

## Downloading by jurisdiction
We can use the `download_network()` functionality to get a vector of stations by jurisdiction. For example, we can choose all the stations in Prince Edward Island using the following:
```{r}
download_realtime(STATION_NUMBER = "ALL", PROV_TERR_STATE_LOC = "PE")
```



### On the distinction between STATIONS() and download_network()
`STATIONS()` and `download_network()` perform similar tasks albeit on different data sources. `STATIONS()` extracts directly from the HYDAT sqlite3 database. In addition to realtime stations, `STATIONS()` outputs discontinued and non-realtime stations:
```{r stations}
STATIONS(STATION_NUMBER = "ALL", PROV_TERR_STATE_LOC = "PE")
```

This is contrast to `download_network()` which downloads all realtime stations. Though this is not always the case, it is best to use `download_network()` when dealing with realtime data and `STATIONS()` when interacting with HYDAT.  

## Example with spatial data
The `download_realtime()` functions allows us to directly query the Environment Canada and Climate Change datamart selecting by station. If we wanted to look at all the realtime stations in a particular hydrologic zone, we could easily do this using the `dplyr`,`sf` and `bcmaps` packages. To install those packages use these commands: 
```{r, echo=TRUE, eval=FALSE}
devtools::install_github("bcgov/bcmaps")
install.packages("sf")
```

And then load these packages. `tidyhydat` and `dplyr` is already loaded above.
```{r, eval = TRUE, echo=TRUE}
library(bcmaps)
library(sf)
```
Now to return the question. BC is divided into hydrologic zones. If use the hydrozones layer in `bcmaps` and convert it to `sf` format, determining which stations reside in which hydrologic zone is trivial. Using `st_join` allows to ask which hydrometric stations in the realtime network (called by `download_network`) are in which hydrologic zones. If we are interested in all realtime stations in the QUEEN CHARLOTTE ISLANDS hydrologic zone, we can generate that list by filtering by the relevant hydrologic zone:
```{r, eval = TRUE, echo=TRUE}
## Convert to sf format
hydrozones_sf <- st_as_sf(bcmaps::hydrozones) %>%
  select(HYDZN_NAME)


qci_stations <- download_network(PROV_TERR_STATE_LOC = "BC") %>%
  st_as_sf(., coords = c("LONGITUDE", "LATITUDE"), 
              crs = 4326, 
              agr = "constant") %>%
  st_transform(crs = 3005) %>%
  st_join(.,hydrozones_sf) %>%
  filter(HYDZN_NAME == "QUEEN CHARLOTTE ISLANDS") %>%
  pull(STATION_NUMBER)

qci_stations
```

Now that vector (`qci_stations`) is useful to select which stations we are interested in. 
```{r}
qci_realtime <- download_realtime(STATION_NUMBER = qci_stations, PROV_TERR_STATE_LOC = "BC")
```

Then using `ggplot2` we could plot these results to have look at the data
```{r}
library(ggplot2)

ggplot(qci_realtime, aes(x = date_time, y = FLOW)) +
  geom_line(aes(colour = STATION_NUMBER))
```

## Project Status

This package is under continual development.

## Getting Help or Reporting an Issue

To report bugs/issues/feature requests, please file an [issue](https://github.com/bcgov/tidyhydat/issues/).

## How to Contribute

If you would like to contribute to the package, please see our 
[CONTRIBUTING](CONTRIBUTING.md) guidelines.

Please note that this project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md). By participating in this project you agree to abide by its terms.

## License

    Copyright 2015 Province of British Columbia

    Licensed under the Apache License, Version 2.0 (the "License");
    you may not use this file except in compliance with the License.
    You may obtain a copy of the License at 

       http://www.apache.org/licenses/LICENSE-2.0

    Unless required by applicable law or agreed to in writing, software
    distributed under the License is distributed on an "AS IS" BASIS,
    WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
    See the License for the specific language governing permissions and
    limitations under the License.
