---
title: "tidyhydat: Extract and Tidy Canadian Hydrometric Data"
authors:
- affiliation: 1
  name: Sam Albers
date: "`r Sys.Date()`"
output:
  html_document:
    keep_md: yes
  pdf_document: default
bibliography: paper.bib
tags:
- R
- tidy data
- hydrology
- Canada
affiliations:
  index: 1
  name: Hydrology and Hydrometric Programs, Ministry of Environment and Climate Change Strategy, British Columbia Provincial Government
---

> "Tidy datasets are all alike but every messy dataset is messy in its own way - "
@wickham2014tidy

```{r options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, messages = FALSE, fig.width = 8, fig.height = 12)
```

```{r packages, warning=FALSE, message=FALSE, echo = FALSE}
library(tidyverse)
library(tidyhydat)
library(dbplyr)
```

# Introduction
Environment and Climate Change Canada (ECCC) through the Water Survey of Canada (WSC) maintains several national hydrometric data sources. These data are partially funded by provincial partners and constitute the main data products of a national integrated hydrometric network. Historical data are stored in the [HYDAT database](http://collaboration.cmc.ec.gc.ca/cmc/hydrometrics/www/). Real-time data are provided by ECCC through two sources: a web service and a datamart. The web service is a login only service which is faster and contains more data that the datamart. Files are updated to the datamart on an hourly basis though the lag between actual hydrometric measurement and the availability of hydrometric data is more like 2.5 hours. The [datamart](http://dd.weather.gc.ca/hydrometric/) is an open data source and is organized in a directory tree structure by province. The objective of this document is the outline the usage of `tidyhydat` [@alberstidyhydat], an R package that accesses these data sources and *tidies* them. The objective of `tidyhydat` is to provide a standard method of accessing ECCC data sources using a consistent and easy to use interface that employs tidy data principles developed by @wickham2014tidy within the R project [@RCore]. 

## Why use R in hydrology?
There are many statistical computing projects that offer great functionality for users. For `tidyhydat` I have chosen to use R. R is a mature open-source project that provides significant potential for advanced modelling, visualization and data manipulation. For hydrologists consdering data analysis tools there are several commonly cited reasons to use R:

- R is and always will be free to use and modify.
- R is easily extensible and comprehensive. It is complimented by a rich suite of packages that implement a vast array of classical and modern statistical methods, exceptionally high-quality graphing capabilities and powerful data manipulation tools to handle a wide variety of data formats.
- R facilitates the scientific method by allowing for a fully reproducible data workflow that can be repeated by others when code is shared.  
- R has a friendly community which is an important infrastructure element of any open source project. 

There have been recent calls to use R more broadly in the field of hydrology [@moore2017watershed]. The `tidyhydat` package is an effort to push this call forward by being a standard package by which hydrologists and other users interact with WSC data in R. Conducting hydrological analysis in a programming environment like R allows hydrologists the ability to create fully reproducible workflows, automate repetitive tasks by letting the computer do the work and provides the same rigour to the data analysis process that hydrologists apply to field equipment and experimental design [@wilson2014best].

## What is tidy data?
Embedded within `tidyhydat` is the principle of *tidy data*. @wickham2014tidy defines tidy data by three principles:

- Each variable forms a column
- Each observation forms a row
- Each type of observational unit forms a table

It is illustrative here to provide an example of the types of data *tidying* processes that `tidyhydat` does for you automatically. The raw `DLY_FLOWS` table in the HYDAT database returns data that looks like this:
```{r, echo = FALSE}
hydat_con <- DBI::dbConnect(RSQLite::SQLite(), system.file("test_db/tinyhydat.sqlite3", package = "tidyhydat"))
tbl(hydat_con, "DLY_FLOWS") 
```

This data structure clearly violates the principles of tidy data - this is messy data. For example, column headers (e.g. `FLOW1`) contain the day number - a value. HYDAT is structured like this for very reasonable historical reasons. It does, however, significantly limit a hydrologists ability to efficiently use hydrometric data. For example, given the current data structure, it is not possible to only extract from the 15th of one month to the 15th of the next. Rather a query would need to be made on all data from the relevant months and then further processing would need to occur.

`tidyhydat` aims to make interacting with HYDAT simpler. For example, we can use the `DLY_FLOWS()` function in `tidyhydat` to query the same data as above but return a much tidier data structure. It is now very simple to extract data between say March 15, 1992 and April 15, 1992. We just need to supply these arguments to `DLY_FLOWS()` after loading the package itself:

```{r, echo = TRUE, message=FALSE}
library(tidyhydat)
DLY_FLOWS(hydat_path = "H:/Hydat.sqlite3",
          STATION_NUMBER = "08MF005",
          start_date = "1992-03-15",
          end_date = "1992-04-15")
```

As you can see, this is much tidier data and is much easier to work with. In addition to these tidy principles, specific to `tidyhydat` we can also define that *for a common data source, variables should be referred to by a common name*. For example, hydrometric stations are given a unique 7 digit identifier that contains important watershed information. This identifier is variously referred to as `STATION_NUMBER` or `ID` depending on the exact ECCC data source. To tidy this hydrometric data, we have renamed where necessary each instance of the unique identifier as `STATION_NUMBER`. This consistency to data formats, and in particular tidy data, situates `tidyhydat` well to interact seamlessly with the powerful tools being developed as the `tidyverse` [@wickham2017tidyverse] and provides a path in R to realizes so of the goals outlined by @moore2017watershed.

# References
